Bayesian spectral density estimation using P-splines with quantile-based knot placement

### 2020-10-27 responses/responses3/response.tex

\response{We would like to thank the referee for a thorough review of our paper. Below, you find a point-by-point response to your points raised. We have now carefully revised the paper taking all of your -- and the other referee's -- comments into account and hope that you will now find the paper acceptable for publication. For your convenience, all changes made are marked in red in the manuscript.}

\textbf{Major comment:}

\refereeQuote{p.8, Quantile-based knot placement\\
I could not understand the meaning of ``calculate its distribution function $F_z = P(z \leq z_i)$" in the second item of the procedure.  The values of the transformed periodogram $z_i$ are regarded as the values of the density $\tilde{f}$, which is non-monotonic, taken at equally-spaced frequencies.  According to item 4, the distribution function $F_z$, and hence the density function $\tilde{f}$, should be defined on the frequency domain. \\
I had a look at the R codes of the package psplinePsd. I think that the density function $\tilde{f}$ is a linear spline that interpolates as
\begin{align*}
\tilde{f} \left(\dfrac{\lambda_k}{2\pi}\right) = \dfrac{z_k}{\sum_{i=1}^{\nu}z_i}, \quad k=1,\dots,\nu.
\end{align*}
}

\response{That's correct!  The $\widetilde{f}$ is actually a probability mass function (not a probability density function), which is interpolated.  However, we have simplified a bit our knot allocation algorithm avoiding that step.  The new version of the algorithm is described step by step in our manuscript.  This change leads to small changes in our application section, which has been consequently updated.  We have also updated the \texttt{psplinePsd} package.}\bigskip

@@

	\response{Fixed.} \bigskip

	\item \refereeQuote{Page 5, line 20, ``If $\tau = \int_0^1f(\tau \omega)d \omega$ is the normalizing constant", it implies $s_r$ is a density, is it correct?}

	\response{Yes, $s_r$ is a pdf.  Note that it is actually a mixture of B-spline densities.}
\end{enumerate}

\subsection*{Further Changes}
Here some extra changes:
\begin{enumerate}
	\item We have defined $K^*$ as the number of internal knots (B-splines and BspCP prior section).
	\item We have specified the prior on $\tau$ (P-spline prior section). 
	\item We have added $s_r(\omega)$ in Eq(3), short notation used in the posterior distribution of~$\tau$ (Posterior computation section).	
	\item We have simplified a bit our knot location approach.  This has led to minor variations in the results shown in the application section, which has been updated.  This change has not impacted the conclusions.  This version of the knot location algorithm is described step by step in the Quantile-based knot placement section.
\end{enumerate}

### 2020-10-27 pSplines.tex
update

{\pmr As shown by \cite{Perron:2001} for distributions on the interval [0,1], both mixtures of B-spline and beta distributions can approximate these arbitrarily well by increasing the number of mixture components, i.e.\ increasing the polynomial order for the beta distributions and the number of knots for suitable know locations for the B-splines. But the rate of approximation is faster for B-splines than for beta distributions.}
%As shown by \cite{Perron:2001} for distributions on the interval [0,1], the mixture of B-spline distributions has better approximation properties than the mixture of beta distributions for a fixed number of components. In particular, the approximation error of the mixture of B-spline distributions can be made arbitrarily small by finding  suitable knot locations and increasing the number of knots {\pmr\citep{Perron:2001}}.
This work motivated \cite{Edwards2019} to propose the BspDP prior on the spectral density function which is based on the representation (\ref{eq:Bernstein}) but with
the beta densities replaced by B-spline densities as defined in the following.

@@

The B-spline basic functions of any order can be recursively defined as
\begin{align*}
B_{k,0}(\omega, \bm{\xi})=&	
\left\{
\begin{array}{ll}
1, & \omega \in [\xi_{k-1}, \xi_{k}] \\
0, & \mbox{otherwise} 
\end{array}
\right.\\
B_{k,r}(\omega, \bm{\xi}) = & \:v_{k,r}B_{k,r-1}(\omega; \bm{\xi}) + \\
& \: (1-v_{k+1,r})B_{k+1,r-1}(\omega; \bm{\xi}),	
\end{align*}
where	   
\begin{align*}	   
v_{k,r}&=	
\left\{
\begin{array}{ll}
\dfrac{\omega - \xi_{k-1}}{\xi_{k+r-1} - \xi_{k-1}}, & \xi_{k-1} \neq \xi_{k+r-1}\\
0, & \mbox{otherwise} 
\end{array}
\right.
\end{align*}
and 
\begin{align*}
\bm{\xi} = \{0 = \xi_0 = \xi_1 = \cdots = \xi_r \leq \xi_{r+1} \leq \cdots \leq \xi_{K}\\ = \xi_{K+1} = \cdots = \xi_{K+r} = 1 \}
\end{align*}
is the knot vector.  This vector is a non-decreasing sequence that contains $K+r+1$ knots, which can be divided in $2r$ external and $K-r+1$ internal knots.  The latter must be $\geq r$.
% external and internal knots as in Edwards et al (2018)

@@

\cite{Eilers2015} consider the use of $K=100$ B-splines a wise choice, unless computational constraints are evident.  Some algorithms that explore different number of knots $K^*$ in a trial sequence and find an optimal value with respect to a certain goodness-of-fit criterion have been proposed. \cite{Ruppert2002} suggested a full-search and myopic algorithms which are based on the generalized cross-validation statistic.  However, their results are not conclusive and this criterion only applies if the penalty parameter is treated as a tuning constant~\citep{Kauermann2011}.  \cite{Likhachev2017} also considered a knot sequence and proposed the selection of the number of knots based on  Akaike's, corrected Akaike's and
the Bayesian Information Criterion.

\cite{Ruppert2002} discussed a heuristic rule of thumb $K^* = \min\left\{n/4, 40\right\}$, which is simpler and, in our experience, seems to work very well in general.  This criterion allocates a knot between $\max\left\{ 4, \lfloor n/40 \rfloor \right\}$ observations, where $\lfloor \cdot \rfloor$ stands for the floor function.  We employ this approach in the application Section for selecting the number of B-spline densities.

@@

{\pmr
The procedure for the frequencies in the unit interval $\omega_l = 2 l/n  \in[0,1]$, which can be easily translated to the frequency domain, can be summarized as follows:	
\begin{enumerate}
	\item Let $\bm{x} = \left (\sqrt{I_n(\omega_1)}, \dots, \sqrt{I_n(\omega_{\nu})} \right)$ be the square root of the periodogram values.
	\item Define $y_l = \left|\dfrac{x_l - \bar{x}}{S_x} \right|$ where $\bar{x}$ denotes the average and $S_x$ the standard deviation of ${\bm x}$.
	\item Define $z_l = \frac{y_l}{\sum_{l=1}^{\nu} y_l}$, and calculate its cumulative distribution $\widetilde{F}$.
	\item Interpolate $\widetilde{F}$.
	\item Let $\bm{\xi}^* = \left(\xi_1^*,\dots,\xi_{K^*}^* \right)$ be a vector of $K^*$ quantiles, equally spaced in $[0,1]$.
	\item The Q-spaced knot vector is given by $\bm{\xi} = \left(\widetilde{F}^{-1}(\xi^*_1),\dots, \widetilde{F}^{-1}(\xi^*_{K^*}) \right)$. %where the inverse function is obtained by interpolation.	
\end{enumerate}

@@

We estimated the spectral density functions using the BspDP and P-spline priors.  The BspDP-based  analysis was performed via the \texttt{gibbs\_bspline} function in the \textsf{R} package \texttt{bsplinePsd} \cite{Edwards:bsplinePsd:2018}.  The algorithm was executed for 100,000 iterations with a burn-in period of 25,000 and thinning factor of 10, resulting in 7,500 samples used for posterior inference.  Pilot analyses showed that these specifications are suitable in these examples in terms of convergence (results not shown).

@@

%REMARK: PLEASE INSERT GRAPH WITH AR(1) AND AR(4) PSD ESTIMATES
\def\arraystretch{1.1}
% IAE
\begin{table}
	\centering
%	\begin{ruledtabular}
		\begin{tabular}{lccc}
			\toprule
			& $n=128$ & $n=256$ & $n=512$ \\ \hline
			AR(1)     &  &  &  \\ 
			B-spline &  0.825 & 0.769 & 0.609  \\
			\textit{Equidistant knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ &  0.769 & 0.655 & 0.527 \\
			\hspace{0.5em}P-spline $d=2$&  0.698 & 0.609 & 0.629 \\
			\textit{Q-spaced knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ & 0.848 & 0.771 & 0.614 \\
			\hspace{0.5em}P-spline $d=2$ & 0.920 & 0.840 & 0.648 \\ \hline
			AR(4)     &  &  &  \\ 
			B-spline & 3.019 & 2.250 & 1.832   \\
			\textit{Equidistant knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ &  2.905 & 2.389 & 2.306\\
			\hspace{0.5em}P-spline $d=2$&  3.149 & 2.566 & 2.387\\
			\textit{Q-spaced knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ & 2.517 & 1.843 & 1.486\\
			\hspace{0.5em}P-spline $d=2$ &  2.750 & 1.989 & 1.673\\
			\bottomrule
		\end{tabular}
%	\end{ruledtabular}
	\caption{Median IAE for the psd estimates using the BspDP and the P-spline priors for simulated AR(1) and AR(4) time series with  different penalty orders ($d$) and knot location schemes for the P-splines.}
	\label{table:sim_IAE}
\end{table}

@@

% Coverage probability Table
\begin{table}
	\centering
%	\begin{ruledtabular}
		\begin{tabular}{lccc}
			\toprule
			& $n=128$ & $n=256$ & $n=512$ \\ \hline
			AR(1)     &  &  &  \\ 
			B-spline &  1.000 & 1.000 & 1.000 \\
			\textit{Equidistant}& & &   \\
			\hspace{0.5em}P-spline $d=1$&  1.000 & 1.000 & 0.997\\
			\hspace{0.5em}P-spline $d=2$& 0.993 & 1.000 & 0.993  \\
			\textit{Q-spaced knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ & 1.000 & 1.000 & 0.997 \\
			\hspace{0.5em}P-spline $d=2$ & 0.933 & 0.527 & 0.713 \\ \hline		       
			AR(4)     &  &  &  \\ 
			B-spline &  0.933 & 0.967 & 0.897  \\
			\textit{Equidistant knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ &  0.547 & 0.343 & 0.010 \\
			\hspace{0.5em}P-spline $d=2$ & 0.320 & 0.170 & 0.007\\
			\textit{Q-spaced knots}& & &   \\	
			\hspace{0.5em}P-spline $d=1$ & 0.800 & 0.833 & 0.803 \\
			\hspace{0.5em}P-spline $d=2$ & 0.313 & 0.363 & 0.277 \\
			\bottomrule
		\end{tabular}
%	\end{ruledtabular}
	\caption{Proportion of times that the true psd is entirely encapsulated within the 90\% uniform credible band of the 300 estimates.}
	\label{table:sim_covPbb}
\end{table}

@@

As \cite{Edwards2019} noted, one of the drawbacks of the BspDP prior  is its computational complexity relative to the BDP.  This is directly reflected in the run-time, which is approximately 2-3 times higher in this example (see Table~3 in \cite{Edwards2019}).  The median run-time in our analyses for the BspDP prior and the P-spline with Q-spaced knots are displayed in Table~\ref{table:sim_time}.  Note that the P-spline method based on equidistant knots has similar run-times to the Q-spaced knots.  The P-spline method is approximately 4-7 times faster than BspDP approach in these examples.  This is due to a reduced number of calculations  per iteration because the B-spline densities are calculated only  once at the start of the algorithm and there are no Dirichlet process approximations via the stick-breaking method.

% TIME Table
\begin{table}
	\centering
%	\begin{ruledtabular}
		\begin{tabular}{lccc}
			\toprule
			& $n=128$ & $n=256$ & $n=512$ \\ \hline
			AR(1)     &  &  &  \\ 
			B-spline &   29.497 & 32.423 & 37.896 \\
			\textit{Q-spaced knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$&  4.859 & 7.450 & 9.606\\
			\hspace{0.5em}P-spline $d=2$&  4.538 & 6.863 & 8.783\\
			B-spline/P-spline &  6.278 & 4.531 & 4.122 \\ \hline
			AR(4)     &  &  & \\ 
			B-spline & 30.111 & 34.126 & 40.875  \\
			\textit{Q-spaced knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ & 4.883 & 7.453 & 9.665 \\
			\hspace{0.5em}P-spline $d=2$& 4.543 & 6.887 & 8.827 \\
			B-spline/P-spline & 6.389 & 4.760 & 4.421 \\	
			\bottomrule
		\end{tabular}
%	\end{ruledtabular}
	\caption{Median run-times in minutes and relative run-times with respect to an average of the P-spline run-times.}
	\label{table:sim_time}
\end{table}

>>>>

{\pmr As shown by \cite{Perron:2001} for distributions on the interval [0,1], both mixtures of B-spline and beta distributions can approximate these arbitrarily well by increasing the number of mixture components, i.e.\ increasing the polynomial order for the beta distributions and the number of knots for suitable knot locations for the B-splines. But the rate of approximation is faster for B-splines than for beta distributions.}
%As shown by \cite{Perron:2001} for distributions on the interval [0,1], the mixture of B-spline distributions has better approximation properties than the mixture of beta distributions for a fixed number of components. In particular, the approximation error of the mixture of B-spline distributions can be made arbitrarily small by finding  suitable knot locations and increasing the number of knots {\pmr\citep{Perron:2001}}.
This work motivated \cite{Edwards2019} to propose the BspDP prior on the spectral density function which is based on the representation (\ref{eq:Bernstein}) but with
the beta densities replaced by B-spline densities as defined in the following.

@@

The B-spline basic functions of any order can be recursively defined as
\begin{align*}
B_{k,0}(\omega, \bm{\xi})=&	
\left\{
\begin{array}{ll}
1, & \omega \in [\xi_{k-1}, \xi_{k}] \\
0, & \mbox{otherwise} 
\end{array}
\right.\\
B_{k,r}(\omega, \bm{\xi}) = & \:v_{k,r}B_{k,r-1}(\omega; \bm{\xi}) + \\
& \: (1-v_{k+1,r})B_{k+1,r-1}(\omega; \bm{\xi}),	
\end{align*}
where	   
\begin{align*}	   
v_{k,r}&=	
\left\{
\begin{array}{ll}
\dfrac{\omega - \xi_{k-1}}{\xi_{k+r-1} - \xi_{k-1}}, & \xi_{k-1} \neq \xi_{k+r-1}\\
0, & \mbox{otherwise} 
\end{array}
\right.
\end{align*}
and 
\begin{align*}
\bm{\xi} = \{0 = \xi_0 = \xi_1 = \cdots = \xi_r \leq \xi_{r+1} \leq \cdots \leq \xi_{K}\\ = \xi_{K+1} = \cdots = \xi_{K+r} = 1 \}
\end{align*}
is the knot vector.  This vector is a non-decreasing sequence that contains $K+r+1$ knots, which can be divided in $2r$ external and ${\pmr K^*}=K-r+1$ internal knots.  The latter must be $\geq r$.
% external and internal knots as in Edwards et al (2018)

@@

\cite{Eilers2015} consider the use of $K=100$ B-splines a wise choice, unless computational constraints are evident.  Some algorithms that explore different number of knots in a trial sequence and find an optimal value with respect to a certain goodness-of-fit criterion have been proposed. \cite{Ruppert2002} suggested a full-search and myopic algorithms which are based on the generalized cross-validation statistic.  However, their results are not conclusive and this criterion only applies if the penalty parameter is treated as a tuning constant~\citep{Kauermann2011}.  \cite{Likhachev2017} also considered a knot sequence and proposed the selection of the number of knots based on  Akaike's, corrected Akaike's and
the Bayesian Information Criterion.

\cite{Ruppert2002} discussed a heuristic rule of thumb $\min\left\{n/4, 40\right\}$, which is simpler and, in our experience, seems to work very well in general.  This criterion allocates a knot between $\max\left\{ 4, \lfloor n/40 \rfloor \right\}$ observations, where $\lfloor \cdot \rfloor$ stands for the floor function.  We employ this approach in the application Section for selecting the number of B-spline densities.

@@

{\pmr
The procedure for the frequencies in the unit interval $\omega_l = 2 l/n  \in[0,1]$, which can be easily translated to the frequency domain, can be summarized as follows:	
\begin{enumerate}
	\item Let $\bm{x} = \left (\sqrt{I_n(\omega_1)}, \dots, \sqrt{I_n(\omega_{\nu})} \right)$ be the square root of the periodogram values.
	\item Define $y_l = \left|\dfrac{x_l - \bar{x}}{S_x} \right|$ where $\bar{x}$ denotes the average and $S_x$ the standard deviation of ${\bm x}$.
	\item Define $z_l = \frac{y_l}{\sum_{l=1}^{\nu} y_l}$, calculate its cumulative distribution $\widetilde{F}$, and interpolate it.
	\item Let $\bm{q} = \left(q_1,\dots,q_{K^*} \right)$ be a vector of $K^*$ quantiles, equally spaced in $[0,1]$.
	\item The Q-spaced internal knot vector is given by $\bm{\xi}^* = \left(\widetilde{F}^{-1}(q_1),\dots, \widetilde{F}^{-1}(q_{K^*}) \right)$. %where the inverse function is obtained by interpolation.	
\end{enumerate}

@@

We estimated the spectral density functions using the BspDP and P-spline priors.  The BspDP-based  analysis was performed via the \texttt{gibbs\_bspline} function in the \textsf{R} package \texttt{bsplinePsd{\pmr 0.6.0}} \citep{Edwards:bsplinePsd:2018}.  The algorithm was executed for 100,000 iterations with a burn-in period of 25,000 and thinning factor of 10, resulting in 7,500 samples used for posterior inference.  Pilot analyses showed that these specifications are suitable in these examples in terms of convergence (results not shown).

@@

%REMARK: PLEASE INSERT GRAPH WITH AR(1) AND AR(4) PSD ESTIMATES
\def\arraystretch{1.1}
% IAE
\begin{table}
	\centering
%	\begin{ruledtabular}
		\begin{tabular}{lccc}
			\toprule
			& $n=128$ & $n=256$ & $n=512$ \\ \hline
			AR(1)     &  &  &  \\ 
			B-spline &  0.870 & 0.714 & 0.594  \\
			\textit{Equidistant knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ &  0.769 & 0.655 & 0.527 \\
			\hspace{0.5em}P-spline $d=2$&  0.698 & 0.609 & 0.629 \\
			\textit{Q-spaced knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ & 0.848 & 0.771 & 0.614 \\
			\hspace{0.5em}P-spline $d=2$ & 0.920 & 0.840 & 0.648 \\ \hline
			AR(4)     &  &  &  \\ 
			B-spline & 2.990 & 2.202 & 1.800   \\
			\textit{Equidistant knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ &  2.905 & 2.389 & 2.306\\
			\hspace{0.5em}P-spline $d=2$&  3.149 & 2.566 & 2.387\\
			\textit{Q-spaced knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ & 2.517 & 1.843 & 1.486\\
			\hspace{0.5em}P-spline $d=2$ &  2.750 & 1.989 & 1.673\\
			\bottomrule
		\end{tabular}
%	\end{ruledtabular}
	\caption{Median IAE for the psd estimates using the BspDP and the P-spline priors for simulated AR(1) and AR(4) time series with  different penalty orders ($d$) and knot location schemes for the P-splines.}
	\label{table:sim_IAE}
\end{table}

@@

% Coverage probability Table
\begin{table}
	\centering
%	\begin{ruledtabular}
		\begin{tabular}{lccc}
			\toprule
			& $n=128$ & $n=256$ & $n=512$ \\ \hline
			AR(1)     &  &  &  \\ 
			B-spline &  1.000 & 1.000 & 0.997 \\
			\textit{Equidistant}& & &   \\
			\hspace{0.5em}P-spline $d=1$&  1.000 & 1.000 & 0.997\\
			\hspace{0.5em}P-spline $d=2$& 0.993 & 1.000 & 0.993  \\
			\textit{Q-spaced knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ & 1.000 & 1.000 & 0.997 \\
			\hspace{0.5em}P-spline $d=2$ & 0.933 & 0.527 & 0.713 \\ \hline		       
			AR(4)     &  &  &  \\ 
			B-spline &  0.907 & 0.970 & 0.897  \\
			\textit{Equidistant knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ &  0.547 & 0.343 & 0.010 \\
			\hspace{0.5em}P-spline $d=2$ & 0.320 & 0.170 & 0.007\\
			\textit{Q-spaced knots}& & &   \\	
			\hspace{0.5em}P-spline $d=1$ & 0.800 & 0.833 & 0.803 \\
			\hspace{0.5em}P-spline $d=2$ & 0.313 & 0.363 & 0.277 \\
			\bottomrule
		\end{tabular}
%	\end{ruledtabular}
	\caption{Proportion of times that the true psd is entirely encapsulated within the 90\% uniform credible band of the 300 estimates.}
	\label{table:sim_covPbb}
\end{table}

@@

As \cite{Edwards2019} noted, one of the drawbacks of the BspDP prior  is its computational complexity relative to the BDP.  This is directly reflected in the run-time, which is approximately 2-3 times higher in this example (see Table~3 in \cite{Edwards2019}).  The median run-time in our analyses for the BspDP prior and the P-spline with Q-spaced knots are displayed in Table~\ref{table:sim_time}.  Note that the P-spline method based on equidistant knots has similar run-times to the Q-spaced knots.  The P-spline method is approximately 4-6 times faster than BspDP approach in these examples.  This is due to a reduced number of calculations  per iteration because the B-spline densities are calculated only  once at the start of the algorithm and there are no Dirichlet process approximations via the stick-breaking method.

% TIME Table
\begin{table}
	\centering
%	\begin{ruledtabular}
		\begin{tabular}{lccc}
			\toprule
			& $n=128$ & $n=256$ & $n=512$ \\ \hline
			AR(1)     &  &  &  \\ 
			B-spline &   24.323 & 28.258 & 34.093 \\
			\textit{Q-spaced knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$&  4.859 & 7.450 & 9.606\\
			\hspace{0.5em}P-spline $d=2$&  4.538 & 6.863 & 8.783\\
			B-spline/P-spline &  6.090 & 4.368 & 4.493 \\ \hline
			AR(4)     &  &  & \\ 
			B-spline & 24.983 & 29.796 & 36.913  \\
			\textit{Q-spaced knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ & 4.883 & 7.453 & 9.665 \\
			\hspace{0.5em}P-spline $d=2$& 4.543 & 6.887 & 8.827 \\
			B-spline/P-spline & 6.062 & 4.591 & 5.523 \\	
			\bottomrule
		\end{tabular}
%	\end{ruledtabular}
	\caption{Median run-times in minutes and relative run-times with respect to an average of the P-spline run-times.}
	\label{table:sim_time}
\end{table}
